---
output: html_document
title: Parallel Processing in PAMpal
layout: default
theme: default
---

```{css, echo=FALSE}
code {
color: #efefef ! important;
background-color: #303030 ! important;
}
pre {
background-color: #303030 ! important;
}
section {
width: 652px;
}
div.sourceCode {
background-color: #303030 ! important;
}
.hljs-literal {
color: #990073;
}
.hljs-number {
color: #099;
}
.hljs-comment {
color: #998;
font-style: italic;
}
.hljs-keyword {
color: #efef8f;
font-weight: bold;
}
.hljs-string {
color: #90a959;
}
```

## Parallel Processing in PAMpal

As of version 1.4.0, `processPgDetections` can benefit from parallel processing
using the `future` package framework. This requires a small amount of setup on 
the user's end, and does have some potential drawbacks, but can greatly reduce
the amount of time required to process really large datasets.

### Parallel Processing Basics

First, a very brief and basic intro on roughly how this works. Most modern
computer processors now come with multiple (usually many!) processing cores,
but by default R will only use one at a time. You may have noticed this if you
have the Task Manager open while processing a large dataset with R, often the
CPU usage will not rise above a fairly low percent even if it sounds like your
computer is working hard. This is because it takes specialized code to make use
of the extra processing power. 

Generally this works by breaking up a larger task into smaller independent pieces
that can be sent off to the different cores to be processed in parallel. For
`PAMpal` this task is loading and applying processing functions to binary file
data. However, an important point is that using multiple cores for processing
requires a certain amount of overhead. New R sessions must be created on the
other cores, and all relevant data and functions have to be passed to those
cores before the processing can happen. There is additional overhead in
managing all the work of sending data back and forth between the cores as they
complete their tasks. So, more cores means faster processing but it also means
more memory and more fixed overhead.

The important thing you need to know for this is that parallel processing
will not be significantly faster for smaller tasks, and using lots of cores
can use up all of your memory.

### Parallel Processing in Practice

Okay, so how do we actually make it happen? Lucky for us, the `future` framework
(specifically the [future.apply][fapply] package) makes it really easy. The premise
of the package is that the end user should always decide what kind of processing to
use, since different users and different environments will have different resources
available. We tell `future` what kind of processing we want to do using the `plan`
function. 

Calling `plan()` will tell you what the current mode is (`sequential` 
means the normal processing you always use).

```{r}
library(future)
plan()
```

To use parallel processing, we use `multisession`, and we need to tell it how
many cores we want to use. `availableCores` will tell us how many our computer
has available.

```{r}
availableCores(logical=FALSE)
```

Then we can set `workers` to some number less than or equal to that. Remember that
more cores means more memory used, so we may not want to use all of them

```{r}
plan(multisession(workers=8))
plan()
```

Now we can just process our data like normal using the regular `PAMpal` functions.
One important note is that **progress bars do not work** when parallel processing,
so you will see it stuck at zero until it finishes. But it should hopefully
finish much faster than before! Also for small to medium projects using
`mode='db'` there is likely only a small difference between parallel and regular
processing due to the overhead incurred, so this is really most valuable for
`mode='recording'` and `mode='time'`.

```{r, eval=FALSE}
data <- processPgDetections(pps, mode='recording')
```

Once you are done, you need to reset your plan back to `sequential` to free up
all the memory that was being used for the multiple-core processing

```{r}
plan(sequential)
```

### TLDR

It works like this:

```{r, eval=FALSE}
library(future)
plan(multisession(workers=min(availableCores(), 8))) # 8 is an arbitrary number I picked
data <- processPgDetections(pps, mode='recording')
plan(sequential)
```

[fapply]: https://www.jottr.org/2018/06/23/future.apply_1.0.0/