---
output: html_document
title: Plotting Call Presence and Density
layout: default
theme: default
---

```{css, echo=FALSE}
code {
color: #efefef ! important;
background-color: #303030 ! important;
}
pre {
background-color: #303030 ! important;
}
div.sourceCode {
background-color: #303030 ! important;
}
.hljs-literal {
color: #990073;
}
.hljs-number {
color: #099;
}
.hljs-comment {
color: #998;
font-style: italic;
}
.hljs-keyword {
color: #efef8f;
font-weight: bold;
}
.hljs-string {
color: #90a959;
}
section {
width: 652px;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load, echo=FALSE, cache=TRUE, include=FALSE}
library(here)
data <- readRDS(here('../Data/BadCCES/ccesStudy.rds'))
```

As of version 1.11.1, the `PAMmisc` package has two new functions to plot
summaries of your detection data, `plotPresBar` and `plotPresGrid`. The
only requirement for these functions is a dataframe with a `UTC` column,
so they can easily be used with `AcousticStudy` objects from `PAMpal`.
It is best if `UTC` is already converted to POSIXct format, but if it is
a character you must provide the date format with the `format` argument
according to the details in `?strptime`.

## Basic Functionality - plotPresBar

First we'll pull out the click detection data from our `AcousticStudy`. This
already has a `UTC` column, so it can immediately be used for the plotting 
functions. The `plotPresBar` function can be used to create a bar graph of
calls across time. The `bin` argument controls the time bins for plotting.
Let's create a plot that shows how many hours of each day have detections
by setting `bin='hours/day'`.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(PAMpal)
clicks <- getClickData(data)
library(PAMmisc)
plotPresBar(clicks, bin='hours/day')
```

We can also show the number of calls per day by changing bin to just
`bin='day'` to show the density of calls rather than just the presence
of calls. This can also be specified as `bin='calls/day'`

```{r}
plotPresBar(clicks, bin='day')
# plotPresBar(clicks, bin='calls/day') # this is the same
```

`bin` can accept time units of `minute`, `hour`, `day`, `week`, and `month`. 
Plural forms of these are also accepted. The
output is a `ggplot` object, so we can also use the `patchwork` package to organize
the plot outputs. 

**NOTE** If the time scale of the bins is too small relative 
to the overall timescale of your data, it is likely that the plot will be incorrect.
The `minutes/hour` plot below is probably missing a lot of bars since there are
more hours than can be reasonably shown across the 4 months of our entire dataset.

```{r}
library(patchwork)
# Using patchwork to organize the plots
(plotPresBar(clicks, bin='days/week') +
    plotPresBar(clicks, bin='hour/week')) /
    (plotPresBar(clicks, bin='minutes/hour') +
         plotPresBar(clicks, bin='days/month'))
```

We can also color the bars by another variable in the dataset using the `by` 
argument to specify the column to color by. Here we show the 
presence of calls of different species by setting `by='species'`. 

**NOTE** This stacks the bars for each species, so it is possible for the
total height of the bar to be greater than 24 hours for a given day. 

```{r}
plotPresBar(clicks, bin='hour/day', by='species')
```

## Basic Functionality - plotPresGrid

The `plotPresGrid` function can be useful to visualize the distribution of
detections across the course of a day. Additionally, if GPS data is present
in the dataframe (or provided separately), it can display dawn/dusk to
investigate diurnal patterns.

The data we've been working with so far cover a range of deployments along
the coast of California, so we'll filter these down to a single deployment
so that the GPS data is for a single location.

```{r, cache=TRUE}
clicks$Drift <- gsub('.*(Drift-[0-9]{1,2}).*', '\\1', basename(clicks$db))
drift4 <- clicks[clicks$Drift == 'Drift-4', ]
```

The data we are working with already has GPS data from earlier in the analysis
(added with `addGps` prior to this tutorial). The gray shaded background marks
the times from dusk to dawn, note that times are always in UTC. With 
`type='presence'`, the blue boxes represent hours that have detections.

```{r}
plotPresGrid(drift4, bin = 'hour', type='presence')
```

With `type='density'` we can see the number of detections in each hour.

```{r}
plotPresGrid(drift4, bin='hour', type='density')
```

For `plotPresGrid`, the `bin` argument can be one of `hour`, `30min`, `15min`, or
`min`.

```{r}
(plotPresGrid(drift4, bin='hour', type='presence') +
     plotPresGrid(drift4, bin='30min', type='presence')) /
    (plotPresGrid(drift4, bin='15min', type='presence') +
         plotPresGrid(drift4, bin='min', type='presence'))
```

Similar to `plotPresBar`, we can also use the `by` argument to create a plot that
contains some kind of label information. For `type='presence'` this will create
boxes of different colors for each label, with a level of transparency controlled
by the `alpha` parameter so that times with multiple labels present can be seen. If
there is a lot of overlap, this can create a messy plot that is hard to read. In this
case the plot can be split into multiple panels using `ggplot2`'s `facet_wrap`.

```{r}
plotPresGrid(drift4, bin='hour', type='presence', by='species')
library(ggplot2)
plotPresGrid(drift4, bin='hour', type='presence', by='species') +
    facet_wrap(vars(species))
```

For `type='density'` using the `by` argument will create a multi-panel plot with `facet_wrap`
since the color is already being used to show the number of calls. You can change the arrangement
of the plot by using `facet_wrap` again and changing the parameters.

```{r}
plotPresGrid(drift4, type='density', by='species')
plotPresGrid(drift4, type='density', by='species') +
    facet_wrap(vars(species), nrow=1)
```

## More Options - Both Functions

Both plotting functions can change the default color of the bars/boxes with the
`fill` argument.

```{r}
plotPresBar(drift4, bin='hour/day', fill='orange') +
    plotPresGrid(drift4, bin='hour', fill='darkgreen')
```

Both plotting functions can add custom titles to the plot with the `title` 
argument.

```{r}
plotPresBar(drift4, bin='hour/day', title='Bar Title') +
    plotPresGrid(drift4, bin='hour', title='Grid Title')
```

Both plotting functions can change the start and end times of the plots with the
`start` and `end` arguments (default is min/max time in the dataset).

```{r}
newStart <- min(drift4$UTC) - 12 * 24 * 3600
newEnd <- max(drift4$UTC) + 7 * 24 * 3600
plotPresBar(drift4, bin='hour/day', start=newStart, end=newEnd) +
    plotPresGrid(drift4, bin='hour', start=newStart, end=newEnd)
```

## More Options - plotPresGrid

The dawn/dusk shading only works if GPS data are provided. In our earlier example,
columns `Longitude` and `Latitude` are already part of the dataframe, so we did 
not need to do anything else. You can also provide GPS coordinates directly using
the `gps` argument if they are not already attached to your data. This should be a
dataframe with columns `UTC`, `Longitude`, and `Latitude`. Here we'll provide coordinates
from a different part of the world to show the difference.

```{r}
newGps <- data.frame(UTC = min(drift4$UTC), Longitude=0, Latitude=56)
plotPresGrid(drift4, bin='hour', title='California') +
    plotPresGrid(drift4, bin='hour', gps=newGps, title='North Sea')
```

In addition to changing the color of the box when plotting presence, we can also change
the box outline color with the `color` argument, and we can change the colormap
used to plot call density with the `cmap` argument.

```{r}
plotPresGrid(drift4, bin='hour', type='presence', color='lightblue', fill='blue') +
    plotPresGrid(drift4, bin='hour', type='density', cmap=viridis::inferno(25))
```